Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = print("world");
let x = print("hello");
...
\end{BVerbatim}
\end{equation}
In this case, substituting a term for a variable changes the order the I/O
actions are performed, which changes the program's observable behaviour. On the
other hand, suppose we have an effect which increments a counter. Then the same
transformation is sound, since \(x\) is used exactly once:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = rand();
let z = rand();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
can change this from a program which always returns true to one which can return
either true or false. On the other hand, if the effect is the possibility of
nontermination, then the substitution is sound. 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure term as one which is all three of
central, copyable, and discardable.

@@

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict}
branching. Our syntax is composed of three main syntactic categories:
\begin{itemize}
  \item \textit{Terms} \(a, b, c, d, e\), which are typed via the judgement
  \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\) is of type
  \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). So the
  centrality \(p\) indicates whether it is safe to substitute this term for a
  variable, and \(q\), plus the linearity of \(A\), control how many times that
  variable can occur: if \(\taff \subseteq q\) then the variable is called
  \textit{affine} or \textit{discardable} and it is permissible to leave it
  unused, whereas if \(\trel \subseteq q\) it is called \textit{relevant} or
  \textit{duplicable} and it is permissible to use multiple times. A term which
  is central, relevant, and affine is called \textit{pure}. The rules for this
  judgement are given in Figure \ref{fig:ssa-term-typing}.

@@

We postulate the existence of a set of \textit{base types} \(X\), as well as a
\textit{unit type} \(\mb{1}\) and \textit{boolean type} \(\mb{2}\). We define
the tensor product of types \(A \otimes B\); we write \(\otimes\) to underscore
the fact that types are allowed to be substructural, assigning every type a
linearity \(\ms{lin}(A)\) satisfying the constraints in Figure
\ref{fig:ssa-structural}. In particular, the rules \rle{unit-lin} and
\rle{bool-lin} state that both the unit and boolean types must be
intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
simply the intersection of the linearity of its components. The rules
\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
the intersection of the linearity of its variables, where the linearity of a
variable is the intersection of the linearity of its type and its quantity
annotation.

@@

\begin{figure}
  \begin{gather*}    
    % \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
    \prftree[r]{\rle{unit-lin}}{}{\tlin{\mathbf{1}}{q}} 
    \qquad
    \prftree[r]{\rle{bool-lin}}{}{\tlin{\mathbf{2}}{q}} 
    \qquad
    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \\
    \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
    \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
      {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
    \prftree[r]{\rle{split-nil}}{\csplits{\cdot}{\cdot}{\cdot}} \qquad
    \prftree[r]{\rle{split-left}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {r \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi}} \qquad
    \prftree[r]{\rle{split-right}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {r \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi, \thyp{x}{A}{r}}} \\
    \prftree[r]{\rle{split-dup}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\ltlin{A}{q}{\trel}}
      {r, s \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi, \thyp{x}{A}{s}}}
      \qquad
    \prftree[r]{\rle{split-drop}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\ltlin{A}{q}{\taff}}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi}}
      \\
    \prftree[r]{\rle{join-nil}}{\lwk{\cdot}{\cdot}} \qquad
    \prftree[r]{\rle{join-cons}}
      {\lwk{\ms{L}}{\ms{K}}}
      {\cwk{\Gamma}{\Delta}}
      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}} 
      \qquad
    \prftree[r]{\rle{join-zero}}
      {\lwk{\ms{L}}{\ms{K}}}
      {\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
  \end{gather*}
  \caption{Structural rules for \isotopessa}
  \Description{Structural rules for isotope-SSA}
  \label{fig:ssa-structural}
\end{figure}

@@

We can now state some of the basic metatheoretic properties of \isotopessa. We
begin with weakening, which we state in our setting as follows:
\begin{lemma}[Weakening] \
  \begin{itemize}
    \item If \(\cwk{\Gamma}{\Delta}\), \(q \subseteq r\), and
    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
    \item If \(\cwk{\Gamma}{\Delta}\) and \(\haslb{\Delta}{t}{\ms{L}}\), then
    \(\haslb{\Gamma}{t}{\ms{L}}\)
    \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{K}}\)
    % \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then
    % \(\lhaslb{\ms{W}}{L}{\ms{K}}\)
  \end{itemize}
\end{lemma}
That is, we can add unused \textit{affine} variables to a context,
\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
remove \textit{affine} variables from existing labels and still keep our typing
judgements valid. This can be proved by a relatively straightforward mutual
induction.

@@

\begin{figure}
  \begin{gather*}
    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
    \\
    \prftree[r]{\rle{subst-nil}}
      {\issubst{\gamma}{\cdot}{\cdot}}
      \qquad
    \prftree[r]{\rle{subst-cons}}
      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
      {\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}
      {\tlin{\Theta_x}{q}}
      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
      {\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
    \\
    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
    \\
    \prftree[r]{\rle{lb-subst-nil}}
    {\lbsubst{\mc{L}}{\cdot}{\ms{K}}}
      \qquad
    \prftree[r]{\rle{lb-subst-cons}}
      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}
      {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
  \end{gather*}
  \caption{ Typing rules for \isotopessa substitutions. We generally assume that
    for all but finitely many variables \(x\) and labels \(\lbl{\ell}\),
    \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\). }
    \Description{Typing rules for isotope-SSA substitutions}
  \label{fig:ssa-subst-typing}
\end{figure}

@@

To solve this problem, we need to introduce the concept of a
\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
that \(\mc{L}(\lbl{\ell}, x)\) is a block that targets \(\ms{K}\). This is
formally defined via the rules \rle{lb-subst-nil}, which states that this is
vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that 
    \begin{itemize}
      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \end{itemize}
  \end{itemize}
\end{lemma}

We can also, using the rules in Figure \ref{fig:ssa-subst-ops}, define
capture-avoiding label-substitution for blocks \([\mc{L}]t\) and CFGs
\([\mc{L}]L\), allowing us to state the following lemma:
\begin{lemma}[Label Substitution] 
  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
  \begin{itemize}
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\),
    \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\),
    \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
  \end{itemize}
\end{lemma}

@@

Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for any
morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
\varnothing\), we have \( (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a)
= \{()\} \). We will see that this corresponds to the fact that \( \ms{let}\;x =
\ms{loop}(); e \) is not semantically equal to \(e\) even where \(x \notin
\ms{fv}(e)\). Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\),
since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A,
|R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
\begin{equation}
  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
\end{equation}
since only the left-hand side contains the pair \((b_0, b_1)\), this
corresponding to the fact that the program \(\ms{let}\;x = \ms{rand}(); (x, x)\) is
not semantically equal to \((\ms{rand}(), \ms{rand}())\).

@@

  \ms{emp}))\) and \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for

>>>>

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = print("world");
let x = print("hello");
...
\end{BVerbatim}
\end{equation}
In this case, substituting a term for a variable changes the order the I/O
actions are performed, which changes the program's observable behaviour. On the
other hand, suppose we have an effect which increments a counter. Then the same
transformation is sound, since \(x\) is used exactly once:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = nondet();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = nondet();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = nondet();
let z = nondet();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{nondet()}, and
can change this from a program which always returns true to one which can return
either true or false. On the other hand, if the effect is the possibility of
nontermination, then the substitution is sound. 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure term as one which is all three of
central, copyable, and discardable.

@@

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict}
branching.

We parametrize an \isotopessa dialect by a set of \textit{base types} \(X \in
\mc{T}\) and, for each pair of types \(A, B\), a set of \textit{instructions}
\(f \in \mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and
\textit{quantity} \(q \subseteq \tint\). As described above, a \textit{central}
instruction is one for which execution order does not matter, while the quantity
\(q\) indicates whether it is safe to delete an instruction if its result is
unused (in which case the instruction is \textit{affine}, and \(\taff \subseteq
q\)) or duplicate an instruction used multiple times (in which case the
instruction is \textit{relevant}, and \(\trel \subseteq q\)). An instruction
which is both affine and relevant is called \textit{pure}. In general, the set
of types is generated by base types \(X\), the \textit{unit type} \(\mb{1}\),
the \textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes
B\). We write \(\otimes\) to underscore the fact that types are allowed to be
substructural: every type is assigned a \textit{linearity} \(\ms{lin}(A)
\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
the linearity of a pair is simply the intersection of the linearity of its
components. 
% The rules \rle{nil-lin} and \rle{cons-lin} together say that the
% linearity of a context is the intersection of the linearity of its variables,
% where the linearity of a variable is the intersection of the linearity of its
% type and its quantity annotation.

Our syntax is composed of three main syntactic categories:
\begin{itemize}
  \item \textit{Terms} \(a, b, c, d, e\), which are typed via the judgement
  \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\) is of type
  \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). So the
  centrality \(p\) indicates whether it is safe to substitute this term for a
  variable, and \(q\), plus the linearity of \(A\), control how many times that
  variable can occur: if \(\taff \subseteq q\) then the variable is called
  \textit{affine} or \textit{discardable} and it is permissible to leave it
  unused, whereas if \(\trel \subseteq q\) it is called \textit{relevant} or
  \textit{duplicable} and it is permissible to use multiple times. A term which
  is central, relevant, and affine is called \textit{pure}. The rules for this
  judgement are given in Figure \ref{fig:ssa-term-typing}.

@@

@@

\begin{figure}
  \begin{gather*}    
    % \prftree[r]{\rle{base-lin}}{q \subseteq \ms{lin}(X)}{\tlin{X}{q}} \qquad
    \prftree[r]{\rle{unit-lin}}{}{\tlin{\mathbf{1}}{q}} 
    \qquad
    \prftree[r]{\rle{bool-lin}}{}{\tlin{\mathbf{2}}{q}} 
    \qquad
    \prftree[r]{\rle{pair-lin}}{\tlin{A}{q}}{\tlin{B}{q}}{\tlin{A \otimes B}{q}} \\
    % \prftree[r]{\rle{nil-lin}}{\tlin{\cdot}{q}} \qquad
    % \prftree[r]{\rle{cons-lin}}{\ltlin{A}{r}{q}}{\tlin{\Gamma}{q}}
    %   {\tlin{\Gamma, \thyp{x}{A}{r}}{q}} \\
    \prftree[r]{\rle{split-nil}}{\csplits{\cdot}{\cdot}{\cdot}} \qquad
    \prftree[r]{\rle{split-left}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {r \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi}} \qquad
    \prftree[r]{\rle{split-right}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {r \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi, \thyp{x}{A}{r}}} \\
    \prftree[r]{\rle{split-dup}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\ltlin{A}{q}{\trel}}
      {r, s \subseteq q}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta, \thyp{x}{A}{r}}{\Xi, \thyp{x}{A}{s}}}
      \qquad
    \prftree[r]{\rle{split-drop}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\ltlin{A}{q}{\taff}}
      {\csplits{\Gamma, \thyp{x}{A}{q}}{\Delta}{\Xi}}
      \\
    \prftree[r]{\rle{join-nil}}{\lwk{\cdot}{\cdot}} \qquad
    \prftree[r]{\rle{join-cons}}
      {\lwk{\ms{L}}{\ms{K}}}
      {\cwk{\Gamma}{\Delta}}
      {\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}} 
      \qquad
    \prftree[r]{\rle{join-zero}}
      {\lwk{\ms{L}}{\ms{K}}}
      {\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}} 
  \end{gather*}
  \caption{Structural rules for \isotopessa}
  \Description{Structural rules for isotope-SSA}
  \label{fig:ssa-structural}
\end{figure}

@@

We can now state some of the basic metatheoretic properties of \isotopessa. We
begin with weakening, which states that we can add unused \textit{affine} variables to a context,
\textit{reduce} the quantity of an term's type, add arbitrary unused labels, and
remove \textit{affine} variables from existing labels and still keep our typing
judgements valid, i.e.
\begin{lemma}[Weakening] \
  \begin{itemize}
    \item If \(\cwk{\Gamma}{\Delta}\), \(q \subseteq r\), and
    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
    \item If \(\cwk{\Gamma}{\Delta}\) and \(\haslb{\Delta}{t}{\ms{L}}\), then
    \(\haslb{\Gamma}{t}{\ms{L}}\)
    \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{t}{\ms{K}}\)
    % \item If \(\lwk{\ms{L}}{\ms{K}}\) and \(\lhaslb{\ms{W}}{L}{\ms{L}}\) then
    % \(\lhaslb{\ms{W}}{L}{\ms{K}}\)
  \end{itemize}
\end{lemma}

@@

\begin{figure}
  \begin{gather*}
    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
    \\
    \prftree[r]{\rle{subst-nil}}
      {\issubst{\gamma}{\cdot}{\cdot}}
      \qquad
    \prftree[r]{\rle{subst-cons}}
      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
      {\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}
      {\tlin{\Theta_x}{q}}
      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
      {\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
    \\
    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
    \\
    \prftree[r]{\rle{lb-subst-nil}}
    {\lbsubst{\mc{L}}{\cdot}{\ms{K}}}
      \qquad
    \prftree[r]{\rle{lb-subst-cons}}
      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}
      {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
  \end{gather*}
  \caption{ Typing rules for \isotopessa substitutions. The linearity of a
    context is defined recursively to be the intersection of the linearities and
    quantities of its variables: \(\ms{lin}(\cdot) = \tint\),
    \(\ms{lin}(\Gamma, \thyp{x}{A}{q}) = \ms{lin}(\Gamma) \cap \ms{lin}(A) \cap
    q\). 
    % We generally assume that for all but finitely many variables \(x\) and
    % labels \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) =
    % \lbrb{\ell}{x}\). 
  } 
  \Description{Typing rules for isotope-SSA substitutions}
  \label{fig:ssa-subst-typing}
\end{figure}

@@

To solve this problem, we need to introduce the concept of a
\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
that \(\mc{L}(\lbl{\ell}, x)\) is a block that targets \(\ms{K}\). This is
formally defined via the rules \rle{lb-subst-nil}, which states that this is
vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that 
    \begin{itemize}
      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \end{itemize}
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), we have that for all
  \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
\end{lemma}

@@

Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for any
morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
\varnothing\), we have \( (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a)
= \{()\} \). We will see that this corresponds to the fact that \( \ms{let}\;x =
\ms{loop}(); e \) is not semantically equal to \(e\) even where \(x \notin
\ms{fv}(e)\). Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\),
since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A,
|R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
\begin{equation}
  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
\end{equation}
since only the left-hand side contains the pair \((b_0, b_1)\), this
corresponding to the fact that the program \(\ms{let}\;x = \ms{nondet}(); (x, x)\) is
not semantically equal to \((\ms{nondet}(), \ms{nondet}())\).

@@

  \ms{emp}))\) and \(\ms{nondet}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for
